---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fibroquant

`fibroquant` quantifies lung fibrosis from histology whole-slide images without 
manual annotation. It reads a slide, optionally splits a multi-section slide 
into its separate sections, and scores tissue with a choice of unsupervised 
analyzers: color clustering, collagen proportionate area, tissue density, and 
texture.

## Reading and splitting a slide

```{r path}
# library(fibroquant)
devtools::load_all()

# vsi <- "/path/to/vsi/image"
vsi <- "/Volumes/Will/Mouse lung 6.10.26/Image_3470.vsi"
has_vsi <- file.exists(vsi)
```

A slide file exposes several image series, each at a handful of resolution
levels. `fq_info()` lists them with pixel dimensions and effective µm/px —
the table `fq_read()` consults to pick the scan series and a working resolution:

```{r info, eval = has_vsi}
fq_info(vsi)
```

`fq_read()` returns an `fq_slide`: an RGB array in `[0, 1]` with its physical
scale and provenance. Printing one reports its dimensions, resolution, and
source:

```{r read, eval = has_vsi}
slide <-
  fq_read(
    vsi,
    target_um_px = 4
  )
slide
```

A slide often carries more than one tissue section. `fq_split()` finds them and
returns an `fq_sections` collection — a list of `fq_section`s that prints a
one-line summary per section:

```{r split, eval = has_vsi}
sections <-
  fq_split(
    slide,
    n = 2
  )
sections
```

## Visualizing

`plot()` dispatches on each object. A slide plots as the whole scan:

```{r plot-slide, eval = has_vsi}
plot(slide)
```

Plotting the collection lays the sections out as a contact sheet:

```{r plot-sheet, eval = has_vsi}
plot(sections)
```

Handing the sections back to the slide draws each crop rectangle on the parent —
the quickest check that the split caught every section and skipped streaks and
debris:

```{r plot-boxes, eval = has_vsi}
plot(slide, sections = sections)
```

A single section plots on its own, cropped tight to its tissue:

```{r plot-section, eval = has_vsi}
plot(sections[[1]])
```

## Scoring fibrosis

Each section is scored by one or more unsupervised analyzers — color clustering,
collagen proportionate area, tissue density, and texture. The workflow fits an
analyzer once on the pooled sections, then reads back a per-section score table
and a severity map for each section. That layer is the next piece under
construction; its chunks slot in here as the analyzers land.

### k-means colour analyzer

The first analyzer reproduces the LungDamage colour approach — unsupervised
CIELAB clustering of tissue colour, clusters ranked dark-to-light into three
severity grades (severe / moderate / healthy) — with two changes. It clusters
only masked tissue pixels: the mask does the job LungDamage's brightest,
background cluster did, so three clusters suffice where MATLAB used four. And it
fits one shared cluster basis across all sections at once (fit-once / apply-many)
instead of re-clustering each image.

```{r kmeans-setup, eval = has_vsi}
slide    <- fq_read(vsi)
sections <- fq_split(slide, n = 2)
sec      <- sections[[1]]
```

#### 1. The spec

`fq_kmeans()` is just the recipe — how many grades, which colour channels, the
blur, the number of k-means restarts. It holds no fitted state.

```{r kmeans-spec, eval = has_vsi}
spec <- 
  fq_kmeans(
    k = 3,
    channels = c("a", "b"),
    smooth_sigma = 2,
    nstart = 3
  )
spec
```

#### 2. Smoothing

A Gaussian blur (sigma in pixels) damps single-pixel stain speckle before
clustering. Same H×W×3 array, slightly softened.

```{r kmeans-smooth, eval = has_vsi}
sm <- fibroquant:::.smooth(sec@rgb, sigma = 2)
EBImage::display(EBImage::Image(sm, colormode = "Color"), method = "raster")
```

#### 3. CIELAB — a and b channels

`rgb2lab`, then keep `a*` and `b*`: the colour axes, with lightness (`L*`) set
aside. The two channels are shown below.

```{r kmeans-lab, eval = has_vsi}
lab <- fibroquant:::.lab(sm)
EBImage::display(EBImage::normalize(lab[, , 2]), method = "raster") # a*
EBImage::display(EBImage::normalize(lab[, , 3]), method = "raster") # b*
```

#### 4. The clustering input

Pull the `a*`/`b*` values at masked tissue pixels only — the clustering input,
and the fibroquant change from LungDamage. An N×2 matrix, one row per tissue
pixel.

```{r kmeans-features, eval = has_vsi}
feat <- fibroquant:::.features(lab, sec@mask, channels = c("a", "b"))
dim(feat)
head(feat)
```

#### 5. Fitting the basis

`fq_fit` pools tissue pixels across every section, runs k-means once, then ranks
the clusters by ascending tissue luminance (`L*`) — darkest grade = most
fibrotic. The fit carries the cluster centres and that severity ordering;
nothing is stored back on the sections.

```{r kmeans-fit, eval = FALSE}
fit <- fq_fit(sections, spec)
fit@centers  # 3 x 2 in a*b* space
fit@severity # cluster -> rank: mean L*, severity 1..3
```

#### 6. Per-section score

`fq_score` applies the fitted basis to one section and measures area per grade.
One row: `severity_index` (area-weighted, 0–1) plus `frac_sev_1` … `frac_sev_3`.

```{r kmeans-score, eval = FALSE}
fq_score(sec, fit)
```

#### 7. The severity map

`fq_render` labels each tissue pixel with its grade, returning an `fq_render`
object (an H×W field, NA off tissue). Plotting it pseudocolours the grades
(severe / moderate / healthy → R/G/B); that plot method lands with the plot
functions.

```{r kmeans-render, eval = FALSE}
rendered <- fq_render(sec, fit)
plot(rendered)
```
