---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fibroquant

`fibroquant` quantifies lung fibrosis from histology whole-slide images without 
manual annotation. It reads a slide, optionally splits a multi-section slide 
into its separate sections, and scores tissue with a choice of unsupervised 
analyzers: color clustering, collagen proportionate area, tissue density, and 
texture.

## Reading and splitting a slide

```{r path}
# library(fibroquant)
devtools::load_all()

# vsi <- "/path/to/vsi/image"
vsi <- "/Volumes/Will/Mouse lung 6.10.26/Image_3470.vsi"
has_vsi <- file.exists(vsi)
```

A slide file exposes several image series, each at a handful of resolution
levels. `fq_info()` lists them with pixel dimensions and effective µm/px —
the table `fq_read()` consults to pick the scan series and a working resolution:

```{r info, eval = has_vsi}
fq_info(vsi)
```

`fq_read()` returns an `fq_slide`: an RGB array in `[0, 1]` with its physical
scale and provenance. Printing one reports its dimensions, resolution, and
source:

```{r read, eval = has_vsi}
slide <-
  fq_read(
    vsi,
    target_um_px = 4
  )
slide
```

A slide often carries more than one tissue section. `fq_split()` finds them and
returns an `fq_sections` collection — a list of `fq_section`s that prints a
one-line summary per section:

```{r split, eval = has_vsi}
sections <-
  fq_split(
    slide,
    n = 2
  )
sections
```

## Visualizing

`plot()` dispatches on each object. A slide plots as the whole scan:

```{r plot-slide, eval = has_vsi}
plot(slide)
```

Plotting the collection lays the sections out as a contact sheet:

```{r plot-sheet, eval = has_vsi}
plot(sections)
```

Handing the sections back to the slide draws each crop rectangle on the parent —
the quickest check that the split caught every section and skipped streaks and
debris:

```{r plot-boxes, eval = has_vsi}
plot(slide, sections = sections)
```

A single section plots on its own, cropped tight to its tissue:

```{r plot-section, eval = has_vsi}
plot(sections[[1]])
```

## Scoring fibrosis

Each section is scored by one or more unsupervised analyzers — color clustering,
collagen proportionate area, tissue density, and texture. The workflow fits an
analyzer once on the pooled sections, then reads back a per-section score table
and a severity map for each section. That layer is the next piece under
construction; its chunks slot in here as the analyzers land.
