---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fibroquant

`fibroquant` quantifies lung fibrosis from Olympus `.vsi` whole-slide images
(bleomycin mouse model; Masson trichrome and H&E) with no manual annotation. It
reads a scan, splits the two lung sections side by side on each slide, and scores
each section by unsupervised color clustering.

## Reading and splitting a slide

```{r path}
# library(fibroquant)
devtools::load_all()

# vsi <- "/path/to/vsi/image"
vsi <- "/Volumes/Will/Mouse lung 6.10.26/Image_3470.vsi"
has_vsi <- file.exists(vsi)
```

The series-and-resolution table the file exposes:

```{r info, eval = FALSE}
fq_slide_info(vsi)
```

Read the scan series at a working resolution near 4 µm/px:

```{r read, eval = FALSE}
slide <-
  fq_read_slide(
    vsi,
    target_um_px = 4
  )
slide
```

Split the two lung sections:

```{r split, eval = FALSE}
sections <-
  fq_split_sections(
    slide,
    n = 2
  )
```

Visualization and the k-means analyzer slot in below as we build them.
