---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r real-setup, include = FALSE}
library(fibroquant)

# The figures below are knit from a local slide. The .Rprofile sets
# FIBROQUANT_TEST_VSI to an OS-specific slide path; point it at any .vsi to
# rebuild them. Chunks are skipped when the file is absent.
vsi <- Sys.getenv("FIBROQUANT_TEST_VSI")
has_vsi <- nzchar(vsi) && file.exists(vsi)
```

# fibroquant

`fibroquant` quantifies lung fibrosis from histology whole-slide images without
manual annotation. It reads a slide, splits a multi-section slide into its
separate sections, and scores the tissue with an unsupervised analyzer. The
analyzer is pluggable: the first one clusters tissue colour in CIELAB space, and
the interface is built so that new approaches drop in beside it.

Every run returns two things: a tidy table of per-section scores, and the fitted
analyzer needed to regenerate a per-pixel severity map for any section.

## Installation

`fibroquant` depends on the Bioconductor packages EBImage and RBioFormats.
Install those first, then the package from GitHub:

```r
# install.packages("BiocManager")
BiocManager::install(c("EBImage", "RBioFormats"))

# install.packages("pak")
pak::pak("wmoldham/fibroquant")
```

## The pipeline

A run moves through four steps, each with its own object type. `fq_run()` chains
them over a folder of slides; the sections below walk through them one at a time.

| Step       | Function                    | Returns                                      |
|------------|-----------------------------|----------------------------------------------|
| Read       | `fq_read()`                 | an `fq_slide`                                |
| Split      | `fq_split()`                | an `fq_sections` of `fq_section`s            |
| Fit        | `fq_fit()`                  | a fitted analyzer, e.g. `fq_kmeans_analyzer` |
| Score, map | `fq_score()`, `fq_render()` | a one-row score table, an `fq_field`         |

```r
library(fibroquant)
vsi <- "path/to/slide.vsi"
```

## Reading and splitting a slide

A slide file holds several image series, each at a handful of resolution levels.
`fq_info()` lists them with pixel dimensions and effective µm/px. This is the
table `fq_read()` consults to pick the scan series and a working resolution.

```{r info, eval = has_vsi}
fq_info(vsi)
```

`fq_read()` returns an `fq_slide`: an RGB array in `[0, 1]` with its physical
scale and provenance. Printing one reports its dimensions, resolution, and
source.

```{r read, eval = has_vsi}
slide <- fq_read(vsi, target_um_px = 4)
slide
```

A slide often carries more than one tissue section. `fq_split()` finds them by
connected components and returns an `fq_sections` collection, a list of
`fq_section`s ordered left to right. Each section adds an analysis mask, a filled
silhouette, and its crop box to the slide it came from.

```{r split, eval = has_vsi}
sections <- fq_split(slide, n = 2)
sections
```

## Visualizing

`plot()` dispatches on each object. A slide plots as the whole scan.

```{r plot-slide, eval = has_vsi}
plot(slide)
```

Plotting the collection lays the sections out as a contact sheet.

```{r plot-sheet, eval = has_vsi}
plot(sections)
```

Handing the sections back to the slide draws each crop box on the parent. This
is the quickest check that the split caught every section and skipped streaks
and debris.

```{r plot-boxes, eval = has_vsi}
plot(slide, sections = sections)
```

A single section plots on its own, cropped tight to its tissue.

```{r plot-section, eval = has_vsi}
plot(sections[[1]])
```

## Analyzers

An analyzer is a **spec** plus three methods:

- `fq_fit(spec, sections)` learns a shared basis from a batch of sections and
  returns a fitted analyzer.
- `fq_score(fit, section)` returns a one-row table of metrics for a section,
  always including a `severity_index` so scores are comparable across analyzers.
- `fq_render(fit, section)` returns a per-pixel severity field for a section,
  which `plot()` pseudocolours.

The basis is fit once and then frozen, so a grade means the same thing on every
section and every slide. Adding an analyzer means implementing these three
methods for a new spec; reading, splitting, batch scoring, and plotting all work
unchanged (see [Adding an analyzer](#adding-an-analyzer)).

### k-means colour analyzer

The first analyzer reproduces the LungDamage colour approach: unsupervised
CIELAB clustering of tissue colour, with clusters ranked by lightness into
severity grades. It differs in two ways. It clusters only masked tissue pixels,
so the mask does the job the brightest background cluster did in MATLAB and three
grades suffice where four were needed. And it fits one shared cluster basis
across all sections at once, rather than re-clustering each image.

**Preprocessing.** Each section is reduced to the pixels and colours the
clustering sees:

1. **Tissue mask.** An Otsu threshold on luminance separates tissue from bright
   glass and airspace, and the airway and other whitespace lumen are excluded.
   The mask is both the density denominator and the pool of pixels to cluster.
2. **Smoothing.** A Gaussian blur damps stain speckle. It is a normalised
   convolution over the mask, so airspace carries no weight into tissue and
   cannot bleed across the rim.
3. **Colour space.** Pixels are converted to CIELAB and clustered on the `a*`
   and `b*` chroma channels, leaving lightness (`L*`) free to order the grades.

The analysis mask is the heart of the preprocessing. It is carried on every
section, so it can be inspected directly.

```{r kmeans-mask, eval = has_vsi}
sec <- sections[[1]]
EBImage::display(EBImage::Image(sec@mask * 1), method = "raster")
```

**The spec.** `fq_kmeans()` is the recipe alone: how many grades, which colour
channels, the blur, and the number of k-means restarts. It holds no fitted
state.

```{r kmeans-spec, eval = has_vsi}
spec <- fq_kmeans(k = 3, channels = c("a", "b"), smooth_sigma = 2, nstart = 3)
spec
```

**Fitting the basis.** `fq_fit()` pools the masked pixels across every section,
runs k-means once, and ranks the clusters by mean lightness so the darkest grade
is the most fibrotic. The fit carries the cluster centres and that ordering.

```{r kmeans-fit, eval = has_vsi}
fit <- fq_fit(spec, sections)
fit@centers   # one row per grade, in a*b* space
fit@luminance # mean L* per grade, mildest to most severe
```

**Scoring.** `fq_score()` applies the basis to one section and measures the area
in each grade. It returns one row: a `severity_index` weighted by area on a 0–10
scale, plus the area fraction in each grade.

```{r kmeans-score, eval = has_vsi}
fq_score(fit, sec)
```

**Mapping.** `fq_render()` labels each tissue pixel with its grade and returns an
`fq_field`. Plotting it pseudocolours the grades from mild (blue) to severe
(red), with off-tissue pixels left white.

```{r kmeans-render, eval = has_vsi}
field <- fq_render(fit, sec)
plot(field)
```

## Batch processing

Scoring generalises from one slide to a folder of them. `fq_manifest()` discovers
the slides and gives each a stable `slide_id` for joining scores back to files.
Add experimental covariates as further columns and they ride through to the
results.

```{r manifest, eval = has_vsi}
# fq_manifest() scans a folder of slides, e.g. "/path/to/slide/folder".
manifest <-
  fq_manifest(dirname(vsi)) |>
  dplyr::filter(!grepl("_01", slide_id)) |>
  dplyr::filter(stringr::str_detect(slide_id, "3470|3472|3478|3479"))
manifest$treatment <- c("saline", "saline", "bleo", "bleo")
manifest
```

`fq_run()` fits the analyzer once on a representative subsample of the slides,
then splits and scores every slide on that shared basis. It returns the
per-section `scores` table with covariates joined on, and the `fit` itself. By
default it fits on up to 25 slides; pass `stratify` to spread that budget across
a covariate, for example `stratify = "treatment"`, so the basis spans the range
of injury.

```{r batch-run, eval = has_vsi}
result <- fq_run(manifest, fq_kmeans(k = 3), progress = FALSE)
result@scores
```

The result also carries everything needed to redraw any section's severity map
after the fact, without refitting. `fq_map()` replays the read-and-split from the
recipe the run recorded and renders the section through the run's fit, returning
a plottable field in one call:

```{r batch-map, eval = has_vsi, fig.show = "hold"}
plot(fq_map(result, "Image_3478", "A"))

# `[[` is the terse shorthand for the same thing:
#   result[["Image_3470", "A"]]
# Drop the section to get every section of a slide, as a named list of maps.
maps <- fq_map(result, "Image_3470")
```

Parallelism is optional and controlled by the caller. Set `mirai::daemons(n)` in
your session and the work for each slide runs across those daemons; with none set
it runs sequentially. A parallel run needs the package installed, because each
daemon loads `fibroquant` to do the scoring.

## Adding an analyzer

A new analyzer is a spec class, a fitted-analyzer class, and the three methods.
The sketch below is a complete, if trivial, analyzer; swap in real logic for the
`...`.

```r
# 1. A spec: hyperparameters only, no fitted state.
fq_threshold <- S7::new_class(
  "fq_threshold",
  parent = fq_spec,
  properties = list(
    cutoff = S7::new_property(S7::class_numeric, default = 0.5)
  )
)

# 2. A fitted analyzer: the basis fq_fit() learns.
fq_threshold_analyzer <- S7::new_class(
  "fq_threshold_analyzer",
  parent = fq_analyzer,
  properties = list(cutoff = S7::class_numeric)
)

# 3. The three methods.
S7::method(fq_fit, fq_threshold) <- function(spec, sections, ...) {
  fq_threshold_analyzer(cutoff = spec@cutoff) # nothing to learn here
}

S7::method(fq_score, fq_threshold_analyzer) <- function(fit, section, ...) {
  tibble::tibble(severity_index = ...) # one row, always a severity_index
}

S7::method(fq_render, fq_threshold_analyzer) <- function(fit, section, ...) {
  fq_field(values = ..., k = ...) # per-pixel grades for plotting
}
```

With those in place, `fq_threshold()` works everywhere `fq_kmeans()` does,
including inside `fq_run()`.
