---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fibroquant

`fibroquant` quantifies lung fibrosis from histology whole-slide images without 
manual annotation. It reads a slide, optionally splits a multi-section slide 
into its separate sections, and scores tissue with a choice of unsupervised 
analyzers: color clustering, collagen proportionate area, tissue density, and 
texture.

## Reading and splitting a slide

```{r path}
# library(fibroquant)
devtools::load_all()

# vsi <- "/path/to/vsi/image"
vsi <- "/Volumes/Will/Mouse lung 6.10.26/Image_3470.vsi"
has_vsi <- file.exists(vsi)
```

List the file's image series and resolution levels with their pixel dimensions 
and effective µm/px — the table `fq_read()` reads to pick the scan series 
and a working resolution:

```{r info, eval = has_vsi}
fq_slide_info(vsi)
```

Read the scan series at a working resolution near 4 µm/px:

```{r read, eval = has_vsi}
slide <-
  fq_read(
    vsi,
    target_um_px = 4
  )
slide
```

Optionally, split a multi-section slide into its sections:

```{r split, eval = has_vsi}
sections <-
  fq_split(
    slide,
    n = 2
  )
sections
```

Visualization and the k-means analyzer slot in below as we build them.
